\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-09-19 doc/paper-jss/sdmTMB-paper.Rnw
Edits

where the expected value $\mathbb{E}[.]$ of an observation $y$ at coordinates in space $\bm{s}$ is equal to mean $\mu_s$. That mean is the result of an inverse link function $g^{-1}$ applied to a linear predictor $\eta$. In this case, that linear predictor is the combination of a model matrix $\bm{X}_{\bm{s}}$ multiplied by a vector of coefficients $\bm{\beta}$ and a value from a spatial GRF $\omega_{\bm{s}}$. This spatial GRF represents the effect of latent spatial variables that are not otherwise accounted for in the model. Alternatively, $\omega_{\bm{s}}$ can be thought of representing spatially correlated noise. More simply, $\omega_{\bm{s}}$ represents a ``wiggly'' surface with an expected value of zero that is added to the prediction in link space.

The vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) is assumed drawn from a multivariate normal distribution with covariance matrix $\boldsymbol{\Sigma}_\omega$ constrained by some covariance function that defines the rate that spatial covariance decays with
distance.
Various covariance functions are possible, but a popular and flexible function is the Mat\'ern (REF) (FIGURE TODO).
The Mat\'ern function describes the covariance $\Phi \left( s_j, s_k \right)$ between spatial locations $s_j$ and $s_k$. If we define $\lVert d \rVert$ as the Euclidean distance between $s_j$ and $s_k$, we can define the covariance as
\[
\Phi\left( s_j, s_k \right) = \frac{\tau^2}{2^{\nu - 1}\Gamma(\nu)}
\left(\kappa \lVert d \rVert \right)^\nu K_\nu \left( \kappa \lVert d \rVert \right),
\]
where $\tau^2$ controls the spatial variance, $\Gamma$ represents the Gamma
function, $K_\nu$ represents the modified Bessel function of the second kind,
and $\kappa$ represents the spatial decorrelation rate. The parameter $\nu$
controls the smoothness of the covariance function, which is fixed at $\nu = 1$, to take advantage of the SPDE \citep{lindgren2011}.
Internally, the parameters $\kappa$ and $\tau$ are converted to correlation range as $\textrm{range} = \sqrt{8} / \kappa$ and to a marginal standard deviation as $\sigma = 1 / \sqrt{4 \pi \exp \left(2 \log(\tau) + 2 \log(\kappa) \right) }$.
The range is the distance that correlation decays to $\approx 0.13$.

The assumption that correlation decays equally in all directions can be relaxed to allow for geometric anisotropy
\[
\Phi\left( s_j, s_k \right) = \frac{\tau^2}{2^{\nu - 1}\Gamma(\nu)}
\left(\kappa \lVert d \rVert \bm{H} \right)^\nu K_\nu \left( \kappa \lVert d \rVert \bm{H} \right).
\]
Here, we add a linear transformation matrix $\bm{H}$ with two estimated parameters governing the major axis direction of geometric anisotropy and the ratio of the major and minor axes \citep{lindgren2011, thorson2015ices, thorson2018comp}. Again, the smoothness parameter $\nu$ is fixed at 1.

@@

We can extend our model to accommodate spatiotemporal data by adding Gaussian random fields for each time step $t$, $\epsilon_{\bm{s},t}$

\[
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}_{\bm{s},t} \bm{\beta} + \omega_{\bm{s}} + \epsilon_{\bm{s},t} \right),\\
\boldsymbol{\omega} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),\\
\boldsymbol{\epsilon_{t}} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right).
\end{aligned}
\]

\subsection{Adding additional model components}

In practice, the above models can become considerably more complex by, for example, structuring the spatiotemporal random fields as autoregressive, letting coefficients vary though time, letting coefficients vary through space as random fields (i.e., SVC REF), or including offsets (REF). All these components are additive in link space. Adopting the notation ``main'' for main effects, ``tvc'' for time-varying coefficients, and ``svc'' for spatially varying coefficients, a maximally complex model could be written as
\[
\begin{aligned}
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma_t} +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
\alpha_g +
O_{\bm{s},t} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} \right).
\end{aligned}
\]
where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector of coefficients that are allowed to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a GRF, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$, $O$ represents an offset variable (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before. We demonstrate all of these model components in the examples later in this paper.

@@

>>>>

where the expected value $\mathbb{E}[.]$ of an observation $y$ at coordinates in space $\bm{s}$ is equal to mean $\mu_s$. That mean is the result of an inverse link function $g^{-1}$ applied to a linear predictor $\eta$. In this case, that linear predictor is the combination of a model matrix $\bm{X}_{\bm{s}}$ multiplied by a vector of coefficients $\bm{\beta}$ and a value from a spatial GRF $\omega_{\bm{s}}$. This spatial GRF represents the effect of latent spatial variables that are not otherwise accounted for in the model. Alternatively, $\omega_{\bm{s}}$ can be thought of representing spatially correlated noise. More simply, $\omega_{\bm{s}}$ represents a ``wiggly'' surface with an expected value of zero that is added to the expectation in link space.

The vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) is assumed drawn from a multivariate normal distribution with covariance matrix $\boldsymbol{\Sigma}_\omega$ constrained by some covariance function that defines the rate that spatial covariance decays with
distance.
Various covariance functions are possible, but a popular and flexible function is the Mat\'ern (REF) (FIGURE TODO).
If we define $\lVert d \rVert$ as the Euclidean distance between spatial points $\bm{s}_j$ and $\bm{s}_k$, we can represent the Mat\'ern covariance $\Phi$ as
\[
\Phi\left( \bm{s}_j, \bm{s}_k \right) = \frac{\tau^2}{2^{\nu - 1}\Gamma(\nu)}
\left(\kappa \lVert d \rVert \right)^\nu K_\nu \left( \kappa \lVert d \rVert \right).
\]
The parameter $\tau^2$ controls the spatial variance, $\Gamma$ represents the Gamma
function, $K_\nu$ represents the modified Bessel function of the second kind,
and $\kappa$ represents the spatial decorrelation rate.
The parameter $\nu$
controls the smoothness of the covariance function, but can be fixed at $\nu = 1$ to take advantage of the SPDE \citep{lindgren2011}.
The parameters $\kappa$ and $\tau$ can be converted to two useful values: (1) correlation range as $\textrm{range} = \sqrt{8} / \kappa$ and (2) marginal standard deviation as $\sigma = 1 / \sqrt{4 \pi \exp \left(2 \log(\tau) + 2 \log(\kappa) \right) }$. The range is the distance that correlation decays to $\approx 0.13$.

The assumption that correlation decays equally in all directions can be relaxed to allow for geometric anisotropy
\[
\Phi\left( s_j, s_k \right) = \frac{\tau^2}{2^{\nu - 1}\Gamma(\nu)}
\left(\kappa \lVert d \rVert \bm{H} \right)^\nu K_\nu \left( \kappa \lVert d \rVert \bm{H} \right).
\]
Here, we add a linear transformation matrix $\bm{H}$ with two estimated parameters governing the major axis direction of geometric anisotropy and the ratio of the major and minor axes \citep{lindgren2011, thorson2015ices, thorson2018comp}.

@@

We can extend our model to accommodate spatiotemporal data by adding Gaussian random fields for each time step $t$, $\bm{\epsilon}_{t}$

\[
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}_{\bm{s},t} \bm{\beta} + \omega_{\bm{s}} + \epsilon_{\bm{s},t} \right),\\
\bm{\omega} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_\omega \right),\\
\bm{\epsilon}_{t} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_{\epsilon} \right).
\end{aligned}
\]

In this case we assume the spatiotemporal random fields are independent at each time step, but we could alternatively assume they are structured as a random walk or autoregressive process (demonstrated in examples below). The spatiotemporal random fields represent latent variables causing spatial correlation that changes with each time step.

\subsection{Additional model components}

In practice, the above models can become considerably more complex by, for example, letting coefficients vary though time, letting coefficients vary through space \citep{hastie1993} as random fields, or including offsets \citep[][p.~206]{mccullagh1989}. All these components are additive in link space. Adopting the notation ``main'' for main effects, ``tvc'' for time-varying coefficients, and ``svc'' for spatially varying coefficients, a maximally complex model could be written as
\[
\begin{aligned}
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma_t} +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
\alpha_g +
O_{\bm{s},t} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} \right).
\end{aligned}
\]
where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector of coefficients that are allowed to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a GRF, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O$ represents an offset variable (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}). We demonstrate these model components in the examples later in this paper.

@@

TODO add paragraph here on uniting functionality not available in any one package?

% TODO: B9: I kept thinking about potential overlaps with spatial / spatiotemporal econometrics. Maybe a short note on how those methods relate to spatial econometrics (e.g. Elhorst, 2012; Lee & Yu, 2010) may broaden the interest in the paper.
